---
title: "Processing datras data for model fitting"
author: "Max Lindmark"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    page-layout: full
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: false
---

```{r}
#| warning: false
#| message: false

library(tibble)
library(tidyr)
library(dplyr)
library(readr)
library(stringr)
library(purrr)

home <- here::here()
```

```{r}
#| warning: false
#| message: false

# Load data from DATRAS sent to me as RData, one for each species.
# I believe some pre-processing was done to these data, I think by Klaas Sys <klaas.sys@ilvo.vlaanderen.be> 
```

### DATRAS data

Below is the function I use to iterate across the .Rdata files, and prepare it to a model-friendly summarize data frame ready for model fitting.

The reason we are working with kg instead of numbers per hour is that it gives us more flexibility when it comes to model distributions and thereby better residuals.

```{r}
# Function to process data to get standardized response variable (kg/h) for a given species and cutoff size
prepare_species_data <- function(file, threshold, depth_mean, depth_sd, ...) {
  
  # Load the specific species Rdata file
  load(paste0(home, "/data/datras_data_cod_had_whg_pok/", file, ".RData"))

  # Estimate length-weight parameters from CA data by year and survey
  lw <- tibble(a = 0.01, b = 3) # In the real script we estimate these by survey*year

  by_size <- dat$HL |>
    bind_cols(lw) |>
    mutate(
      size_class = if_else(LngtCm < threshold, "small", "large"),
      biomass_g = Count * a * LngtCm^b, # <- Key column here! Count is not a DATRAS original column I think, hence I need to create that in the Norwegian data
      biomass_kg = biomass_g / 1000
    ) |>
    summarise(
      n = sum(Count, na.rm = TRUE),
      biomass_kg = sum(biomass_kg, na.rm = TRUE),
      a = first(a),
      b = first(b),
      .by = c(haul.id, size_class)
    )

  d <- expand_grid( # <- Make an empty df with each haul id repeated for each size class
    haul.id = dat$HH$haul.id,
    size_class = c("small", "large")
  ) |> 
    left_join(by_size, by = c("haul.id", "size_class")) |>  # <- Join observed catches by haul id
    mutate(
      n = replace_na(n, 0), # <- If no catches, n will be NA. Replace this with 0
      biomass_kg = replace_na(biomass_kg, 0)
    ) |>
    left_join( # <- Add in HH columns
      dat$HH |> dplyr::select(
        haul.id, Year, Survey, Gear, lon, lat, Depth,
        TimeShotHour, Ship, HaulDur, Month
      ),
      by = "haul.id"
    ) |> 
    janitor::clean_names() |>
    filter(!is.na(haul_dur))
    
  d
}
```

That's the whole processing script to get kg/h for a give size class. Below is a script to run this for a give set of species and threshold sizes.

```{r}
# Species metadata grid with settings
species_grid <- tibble(
  species = c("cod", "had", "whg"),
  file = c("datras_data_cod", "datras_data_had", "datras_data_whg"), # "datras_data_pok"), # skipping saithe because few juveniles caught
  threshold = c(35, 30, 27), # Size thresholds
  cutoff = 15 # for mesh. Minimum distance between mesh verticies
)

# Loop function across all species
dat <- list()

for (i in 1:nrow(species_grid)) {
  tryCatch(
    {
      sp <- species_grid$species[i]
      file <- species_grid$file[i]
      threshold <- species_grid$threshold[i]
      cutoff <- species_grid$cutoff[i]

      dat[[i]] <- prepare_species_data(
        file = file,
        threshold = threshold,
        depth_mean = depth_mean,
        depth_sd = depth_sd
      ) |>
        mutate(
          species = sp,
          threshold = threshold,
          cutoff = cutoff
        )

      cat("Prepared:", sp, "\n")
    },
    error = function(e) {
      cat("ERROR preparing", sp, ":", conditionMessage(e), "\n")
    }
  )
}
```

### Norwegian shrimp survey

Now we want to add the Norwegian shrimp survey data. This is not of class `DATRASraw` but instead a .csv file, where each `RecordType` is just appended. Below is a script that subsets each `RecordType` into a separate object with correct headings.

```{r}
#| warning: false
#| message: false

# Read raw file lines
lines <- read_lines(
  paste0(home, "/data/NO_shrimp_survey_gadoids_DATRAS.csv")
)

# Find all header rows
header_idx <- which(str_detect(lines, "^RecordType"))

# Add artificial end position
header_end <- c(header_idx[-1] - 1, length(lines))

# Function to parse one block
parse_block <- function(start, end) {

  header <- lines[start]

  # Determine record type from first data row
  first_data <- lines[start + 1]
  rec_type <- str_split(first_data, "\t|,", simplify = TRUE)[1]

  # Extract lines for this block
  block_lines <- lines[start:end]

  # Create temporary text connection
  block_text <- paste(block_lines, collapse = "\n")

  # Read as dataframe
  df <- read_delim(
    I(block_text),
    delim = ",",
    show_col_types = FALSE,
    trim_ws = TRUE
  )

  list(
    type = rec_type,
    data = df
  )
}

# Parse all blocks
blocks <- map2(header_idx, header_end, parse_block)

# Combine blocks of same RecordType
no_HH <- blocks |>
  keep(~ .x$type == "HH") |>
  map("data") |>
  bind_rows()

no_HL <- blocks |>
  keep(~ .x$type == "HL") |>
  map("data") |>
  bind_rows()

no_CA <- blocks |>
  keep(~ .x$type == "CA") |>
  map("data") |>
  bind_rows()
```

Now we have something that looks better. I have separate files for each `RecordType` in both the DATRAS data and the Norwegian data, and in theory I should be able to `bind_rows()` them before running `prepare_species_data()`.

### Combine Norwegian shrimp survey data with DATRAS data

```{r}
# Check Norwegian HL data
head(no_HL, 2) |> as.data.frame()

# Check DATRAS HL data
load(paste0(home, "/data/datras_data_cod_had_whg_pok/datras_data_cod.RData"))
head(dat$HL, 2) |> as.data.frame()

head(dat$HH$haul.id, 1) |> droplevels()
head(dat$HH, 2) |> as.data.frame()

# And check the column names in both HL files:
sort(names(dat$HL))
sort(names(no_HL))
```

Now, I can't just add these together because they differ in names and in number of columns. Some are easy; `LengthClass = LngtClas`, for example, but many are not.
ggsave(paste0("figures/survey_map.png"), width = 13, height = 14.5, units = "cm")

### Specific questions for you:

* **Klaas Sys**: Is my approach correct? Is `Count` the column that indicates the total count of fish of that length class caught in that given haul? I.e., is it already raised to the haul if sub-sampled?

* **Klaas Sys**: If so, how was the count column created, so that I can reproduce it with the Norwegian data (I assume it was created in an earlier step because I don't think it is a default column in DATRAS exchange)

* **Norwegian**: Given that column names do not exactly match those of DATRAS exchange data, what would the corresponding calculation be on the shrimp data be to end up with what Klaas had done?


```{r}
hist(dat$HL$Count)

summary(no_HL$NumberAtLength)
summary(dat$HL$HLNoAtLngt)
```


