---
title: "Processing DATRAS and Norwegian data for model fitting"
author: "Max Lindmark"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    page-layout: full
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: false
---

```{r}
#| warning: false
#| message: false

library(concaveman)
library(marmap)
library(forcats)
library(stringr)
library(sdmTMB)
library(tidyr)
library(dplyr)
library(purrr)
library(ggplot2)
library(rnaturalearth)
library(sf)
library(readr)
library(ggsidekick)
theme_set(theme_sleek())
library(viridis)

home <- here::here()

# Convert lat/lon to ETRS89-LAEA (EPSG:3035) projected coordinates to minimize distortion in the domain,
# which is important because the model estimates the range of spatial autocorrelation,
# specifically distance where correlation <0.13
# Change to  them from metres to kilometres to help estimation
crs_lcc <- 3035 # ETRS89-LAEA

# to_lcc <- function(df, lon = "lon", lat = "lat") {
#   df |>
#     sf::st_as_sf(coords = c(lon, lat), crs = 4326) |>
#     sf::st_transform(crs_lcc) |>
#     dplyr::mutate(
#       X = sf::st_coordinates(geometry)[, 1] / 1000,
#       Y = sf::st_coordinates(geometry)[, 2] / 1000
#     ) |>
#     sf::st_drop_geometry()
# }

to_lcc <- function(df, lon = "lon", lat = "lat") {
  df |>
    sf::st_as_sf(
      coords = c(lon, lat),
      crs = 4326,
      remove = FALSE
    ) |>
    sf::st_transform(crs_lcc) |>
    dplyr::mutate(
      X = sf::st_coordinates(geometry)[, 1] / 1000,
      Y = sf::st_coordinates(geometry)[, 2] / 1000
    ) |>
    sf::st_drop_geometry()
}

# For map plots
world <- ne_countries(scale = "medium", returnclass = "sf")
world_lcc <- world |> st_transform(crs = crs_lcc)
```

```{r}
#| warning: false
#| message: false

# Load data from DATRAS sent to me as RData, one for each species.
# Some pre-processing was done to these data, I think by Klaas Sys <klaas.sys@ilvo.vlaanderen.be>
```

### Load Norwegian shrimp survey data
Add the Norwegian shrimp survey data. This is not of class `DATRASraw` but instead a .csv file, where each `RecordType` is just appended. Below is a script that subsets each `RecordType` into a separate object with correct headings.

```{r}
#| warning: false
#| message: false

# Read raw file lines
lines <- read_lines(
  paste0(home, "/data/NO_shrimp_survey_gadoids_DATRAS.csv")
)

# Find all header rows
header_idx <- which(str_detect(lines, "^RecordType"))

# Add artificial end position
header_end <- c(header_idx[-1] - 1, length(lines))

# Function to parse one block
parse_block <- function(start, end) {
  header <- lines[start]

  # Determine record type from first data row
  first_data <- lines[start + 1]
  rec_type <- str_split(first_data, "\t|,", simplify = TRUE)[1]

  # Extract lines for this block
  block_lines <- lines[start:end]

  # Create temporary text connection
  block_text <- paste(block_lines, collapse = "\n")

  # Read as dataframe
  df <- read_delim(
    I(block_text),
    delim = ",",
    show_col_types = FALSE,
    trim_ws = TRUE
  )

  list(
    type = rec_type,
    data = df
  )
}

# Parse all blocks
blocks <- map2(header_idx, header_end, parse_block)

# Combine blocks of same RecordType
no_HH <- blocks |>
  keep(~ .x$type == "HH") |>
  map("data") |>
  bind_rows()

no_HL <- blocks |>
  keep(~ .x$type == "HL") |>
  map("data") |>
  bind_rows() |>
  tidylog::filter(!SpeciesCode == "126441") # Remove saithe

no_CA <- blocks |>
  keep(~ .x$type == "CA") |>
  map("data") |>
  bind_rows() |>
  tidylog::filter(!SpeciesCode == "126441") # Remove saithe
```

### Summarise Norwegian data
Note, this code is somewhat different from the pre-processed DATRAS data (!)

```{r}
#| warning: false
#| message: false

# Species metadata grid with settings
species_thresh <- tibble(
  species = c("cod", "had", "whg"),
  SpeciesCode = c(126436, 126437, 126438),
  threshold = c(35, 30, 27),
)

# Estimate length-weight parameters from CA data by year and survey
lw <- no_CA |>
  rename(
    NoAtLngt = NumberAtLength,
    LngtCm = LengthClass,
    IndWgt = IndividualWeight
  ) |>
  mutate(
    length_cm = LngtCm,
    Quarter = as.factor(Quarter),
    DoorType = as.factor(DoorType),
    Year = as.factor(Year),
    Survey = as.factor(Survey)
  ) |>
  filter(IndWgt > 0, LngtCm > 0) |>
  tidyr::uncount(NoAtLngt) |>
  group_by(Year, Survey) |>
  group_modify(~ broom::tidy(lm(log(IndWgt) ~ log(length_cm), data = .x))) |>
  ungroup() |>
  dplyr::select(Year, Survey, term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  mutate(a = exp(`(Intercept)`), b = `log(length_cm)`) |>
  dplyr::select(Year, Survey, a, b) |>
  janitor::clean_names() |>
  mutate(
    b = if_else(b < 2.9 | b > 3.1, NA, b),
    a = if_else(a < quantile(a, 0.2) | a > quantile(a, 0.8), NA, a)
  )

a_fallback <- median(lw$a, na.rm = TRUE)
b_fallback <- median(lw$b, na.rm = TRUE)

# Add count to HL data. First we need a haul.id so that we can join HH and HL
no_HL <- no_HL |>
  mutate(haul.id = paste(Year, Quarter, Country, "Ship", StationName, HaulNumber, sep = ":"))

no_HH <- no_HH |>
  mutate(haul.id = paste(Year, Quarter, Country, "Ship", StationName, HaulNumber, sep = ":"))

no_HL <- no_HL |>
  left_join(no_HH |>
    distinct(haul.id, .keep_all = TRUE) |>
    dplyr::select(haul.id, DataType, HaulDuration)) |>
  mutate(
    multiplier1 = ifelse(DataType == "C", HaulDuration / 60, 1),
    multiplier2 = ifelse(!is.na(SubsamplingFactor), SubsamplingFactor, 1),
    Count = NumberAtLength * multiplier1 * multiplier2
  )

# Summarise kg per haul, size-class and species
by_size <- no_HL |>
  rename(length_cm = LengthClass) |>
  mutate(
    Year = as.factor(Year),
    LngtCm = length_cm,
    Survey = as.factor(Survey)
  ) |>
  left_join(species_thresh, by = "SpeciesCode") |>
  left_join(lw, by = c("Year" = "year", "Survey" = "survey")) |>
  mutate(
    Count = ifelse(Count < 0, NA, Count),
    a = replace_na(a, a_fallback),
    b = replace_na(b, b_fallback),
    size_class = if_else(LngtCm < threshold, "small", "large"),
    biomass_g = Count * a * LngtCm^b,
    biomass_kg = biomass_g / 1000
  ) |>
  summarise(
    n = sum(Count, na.rm = TRUE),
    biomass_kg = sum(biomass_kg, na.rm = TRUE),
    .by = c(haul.id, size_class, species)
  )

nor_d <- expand_grid(
  haul.id = no_HH$haul.id,
  size_class = c("small", "large"),
  species = unique(by_size$species)
) |>
  left_join(by_size, by = c("haul.id", "size_class", "species")) |>
  mutate(
    n = replace_na(n, 0),
    biomass_kg = replace_na(biomass_kg, 0)
  ) |>
  left_join(
    no_HH |>
      mutate(
        Gear2 = "unspecified bottom trawl",
        Survey = "Norwegian-Shrimp",
        Ship = "Ship"
      ) |>
      rename(
        lat = ShootLatitude,
        lon = ShootLongitude,
        HaulDur = HaulDuration
      ) |>
      dplyr::select(
        haul.id, Year, Survey, lon, lat, Gear2, HaulDur, Ship, Month
      ) |>
      distinct(haul.id, .keep_all = TRUE),
    by = "haul.id"
  ) |>
  rename(Gear = Gear2) |>
  janitor::clean_names()
```

### DATRAS data

Below is the function I use to iterate across the .Rdata files, and prepare it to a model-friendly summarize data frame ready for model fitting.

```{r}
#| warning: false
#| message: false

# Function to process data to get standardized response variable (kg/h) for a given species and cutoff size
prepare_species_data <- function(file, threshold, ...) {
  # Load the specific species Rdata file
  load(paste0(home, "/data/datras_data_cod_had_whg_pok/", file, ".RData"))

  # Estimate length-weight parameters from CA data by year and survey
  lw <- dat$CA |>
    drop_na(IndWgt, LngtClas) |>
    filter(IndWgt > 0, LngtClas > 0) |>
    tidyr::uncount(NoAtALK) |>
    mutate(length_cm = ifelse(LngtCode == ".", LngtClas / 10, LngtClas)) |>
    group_by(Year, Survey) |>
    group_modify(~ broom::tidy(lm(log(IndWgt) ~ log(length_cm), data = .x))) |>
    ungroup() |>
    dplyr::select(Year, Survey, term, estimate) |>
    pivot_wider(names_from = term, values_from = estimate) |>
    mutate(a = exp(`(Intercept)`), b = `log(length_cm)`) |>
    dplyr::select(Year, Survey, a, b) |>
    janitor::clean_names() |>
    mutate(
      b = if_else(b < 2.9 | b > 3.1, NA, b),
      a = if_else(a < quantile(a, 0.2) | a > quantile(a, 0.8), NA, a)
    )

  a_fallback <- median(lw$a, na.rm = TRUE)
  b_fallback <- median(lw$b, na.rm = TRUE)

  by_size <- dat$HL |>
    left_join(lw, by = c("Year" = "year", "Survey" = "survey")) |>
    mutate(
      a = replace_na(a, a_fallback),
      b = replace_na(b, b_fallback),
      size_class = if_else(LngtCm < threshold, "small", "large"),
      biomass_g = Count * a * LngtCm^b,
      biomass_kg = biomass_g / 1000
    ) |>
    summarise(
      n = sum(Count, na.rm = TRUE),
      biomass_kg = sum(biomass_kg, na.rm = TRUE),
      .by = c(haul.id, size_class)
    )

  d <- expand_grid( # <- Make an empty df with each haul id repeated for each size class
    haul.id = dat$HH$haul.id,
    size_class = c("small", "large")
  ) |>
    left_join(by_size,
      by = c("haul.id", "size_class")
    ) |> # <- Join observed catches by haul id
    mutate(
      n = replace_na(n, 0), # <- If no catches, n will be NA. Replace this with 0
      biomass_kg = replace_na(biomass_kg, 0)
    ) |>
    left_join( # <- Add in HH columns
      dat$HH |>
        dplyr::select(
          haul.id, Year, Survey, Gear, lon, lat,
          Ship, HaulDur, Month
        ),
      by = "haul.id"
    ) |>
    janitor::clean_names() |>
    filter(!is.na(haul_dur))

  d
}
```

That's the whole processing script to get kg/h for a give size class. Below is a script to run this for a give set of species and threshold sizes.

```{r}
#| warning: false
#| message: false

# Species metadata grid with settings
species_grid <- tibble(
  species = c("cod", "had", "whg"),
  file = c("datras_data_cod", "datras_data_had", "datras_data_whg"), # "datras_data_pok"), # skipping saithe because few juveniles caught
  threshold = c(35, 30, 27)
) # Size thresholds

# Loop function across all species
dd <- list()

for (i in 1:nrow(species_grid)) {
  tryCatch(
    {
      sp <- species_grid$species[i]
      file <- species_grid$file[i]
      threshold <- species_grid$threshold[i]

      dd[[i]] <- prepare_species_data(
        file = file,
        threshold = threshold
      ) |>
        mutate(
          species = sp
        )

      cat("Prepared:", sp, "\n")
    },
    error = function(e) {
      cat("ERROR preparing", sp, ":", conditionMessage(e), "\n")
    }
  )
}

datras_d <- bind_rows(dd)
```

### Combine Norwegian and DATRAS data

```{r}
#| warning: false
#| message: false

d <- bind_rows(
  datras_d |>
    mutate(year = as.numeric(as.character(year))),
  nor_d |>
    mutate(
      haul_id = as.factor(haul_id),
      month = as.integer(month)
    )
) |>
  filter(year >= 1993) |>
  to_lcc(lon = "lon", lat = "lat") |>
  mutate(
    quarter = ceiling(month / 3),
    sp_sz = paste(species, size_class, sep = "_")
  ) |>
  mutate(
    sp_sz = paste(species, size_class, sep = "_"),
    sp_sz = fct_recode(
      sp_sz,
      "Cod < 35 cm" = "cod_small",
      "Haddock < 30 cm" = "had_small",
      "Whiting < 27 cm" = "whg_small",
      "Cod ≥ 35 cm" = "cod_large",
      "Haddock ≥ 30 cm" = "had_large",
      "Whiting ≥ 27 cm" = "whg_large"
    )
  )
```

### Make a prediction grid

```{r}
#| warning: false
#| message: false

# Convert lat/lon to ETRS89-LAEA (EPSG:3035) projected coordinates to minimize distortion in the domain,
# which is important because the model estimates the range of spatial autocorrelation,
# specifically distance where correlation <0.13
# Change to  them from metres to kilometres to help estimation

# Using NOAA bathy from marmap to get depth raster. Likely there are more regional shapefiles
# that have a higher resolution, but for now this will have to do

# This is a fiddly grid but my normal approach of a convex hull of the sampling data doesn't play well with UK(!)
grid_sf <- d |>
  distinct(X, Y) |>
  st_as_sf(coords = c("X", "Y"), crs = NA)

hull <- concaveman(grid_sf, concavity = 2)
# plot(hull)
hull_buf <- st_buffer(hull, dist = 20)
# plot(hull_buf)
coords <- st_coordinates(hull_buf)[, 1:2]

sp_poly <- sp::SpatialPolygons(
  list(sp::Polygons(list(sp::Polygon(coords)), ID = 1))
)
sp_poly_df <- sp::SpatialPolygonsDataFrame(
  sp_poly,
  data = data.frame(ID = 1)
)

# Size of cell width (in km)
cell_width <- 3

spatial_grid <- expand.grid(
  X = seq(min(d$X), max(d$X), cell_width),
  Y = seq(min(d$Y), max(d$Y), cell_width)
)

sp::coordinates(spatial_grid) <- c("X", "Y")
inside <- !is.na(sp::over(spatial_grid, as(sp_poly_df, "SpatialPolygons")))
spatial_grid <- as.data.frame(spatial_grid[inside, ])

pred_grid <- expand_grid(
  spatial_grid,
  year = sort(unique(d$year))
) |>
  mutate(
    lon = st_coordinates(
      st_transform(
        st_as_sf(data.frame(X = X * 1000, Y = Y * 1000),
          coords = c("X", "Y"), crs = crs_lcc
        ), 4326
      )
    )[, 1],
    lat = st_coordinates(
      st_transform(
        st_as_sf(data.frame(X = X * 1000, Y = Y * 1000),
          coords = c("X", "Y"), crs = crs_lcc
        ), 4326
      )
    )[, 2],
    gear = as.factor("GOV"),
    ship_f = factor(d$ship[1]),
    time_shot_hour = 12,
    haul_dur = 30,
    haul_dur_sc = 0,
    year_f = as.factor(year),
    gear_f = as.factor(gear),
    month = 8,
    month_f = as.factor(month),
    quarter_f = as.factor(3)
  )

# ggplot(pred_grid |> filter(year == 2002), aes(X, Y)) +
#   geom_raster() +
#   geom_point(data = d, aes(X, Y, color = "data"))

# Add depth
depth_box <- getNOAA.bathy(min(pred_grid$lon) - .1, max(pred_grid$lon) + .1, min(pred_grid$lat) - .1, max(pred_grid$lat) + .1)
pred_grid$depth <- get.depth(depth_box, x = pred_grid$lon, y = pred_grid$lat, locator = F)$depth
pred_grid$depth <- pred_grid$depth * (-1)

pred_grid <- pred_grid |>
  filter(depth > 1)

# Scale depth (use that for data also)
depth_mean <- mean(pred_grid$depth, na.rm = TRUE)
depth_sd <- sd(pred_grid$depth, na.rm = TRUE)

pred_grid <- pred_grid |>
  mutate(
    depth_sc = (depth - depth_mean) / depth_sd,
    depth_sc2 = depth_sc^2
  )

# Test plot
world <- ne_countries(scale = "medium", returnclass = "sf")
world_lcc <- world |> st_transform(crs = crs_lcc)

ggplot() +
  geom_raster(
    data = pred_grid |> filter(year == max(year)),
    aes(x = X * 1000, y = Y * 1000, fill = depth)
  ) +
  scale_fill_viridis() +
  geom_sf(data = world_lcc) +
  coord_sf(
    xlim = range(pred_grid$X) * 1000,
    ylim = range(pred_grid$Y) * 1000,
    crs = crs_lcc,
    expand = 0
  ) +
  guides(fill = guide_colorbar(
    title.position = "top", title.hjust = 0.5,
    position = "inside", name = "Depth"
  )) +
  labs(x = "Easting (km)", y = "Northing (km)") +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.direction = "horizontal",
    legend.position.inside = c(0.63, 0.057),
    legend.key.width = unit(1, "cm")
  )
```

### Explore data

```{r}
#| warning: false
#| message: false

data_plot <- ggplot() +
  facet_wrap(~year, ncol = 7) +
  geom_sf(data = world_lcc) +
  coord_sf(
    xlim = range(pred_grid$X),
    ylim = range(pred_grid$Y),
    crs = crs_lcc,
    expand = 0
  ) +
  guides(color = guide_legend(
    title.position = "top", title.hjust = 0.5,
    position = "inside"
  )) +
  labs(x = "Easting (km)", y = "Northing (km)") +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.direction = "horizontal",
    legend.position.inside = c(0.93, 0.08),
    legend.key.width = unit(1, "cm")
  )

ggplot() +
  geom_point(
    data = d |> distinct(haul_id, .keep_all = TRUE),
    aes(X * 1000, Y * 1000, color = as.factor(survey)),
    alpha = 0.25, size = 0.75
  ) +
  geom_sf(data = world_lcc) +
  theme(legend.position = "bottom") +
  coord_sf(
    xlim = range(d$X) * 1000,
    ylim = range(d$Y) * 1000,
    crs = crs_lcc,
    expand = 0
  ) +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.direction = "horizontal",
    legend.position.inside = c(0.96, 0.07),
    legend.title = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(x = "Longitude", y = "Latitude") +
  scale_color_brewer(palette = "Dark2", direction = -1)

ggsave(paste0(home, "/figures/data/survey_map.png"), width = 13, height = 14.5, units = "cm")

# Plot data by month/quarter
ggplot() +
  geom_point(
    data = d |> distinct(haul_id, .keep_all = TRUE),
    aes(X * 1000, Y * 1000, color = as.factor(survey)),
    alpha = 0.25, size = 0.75
  ) +
  geom_sf(data = world_lcc) +
  facet_wrap(~month) +
  theme(legend.position = "bottom") +
  coord_sf(
    xlim = range(d$X) * 1000,
    ylim = range(d$Y) * 1000,
    crs = crs_lcc,
    expand = 0
  ) +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.direction = "horizontal",
    legend.position.inside = c(0.96, 0.07),
    legend.title = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(x = "Longitude", y = "Latitude") +
  scale_color_brewer(palette = "Dark2", direction = -1)


ggplot() +
  geom_point(
    data = d |> distinct(haul_id, .keep_all = TRUE),
    aes(X * 1000, Y * 1000, color = as.factor(survey)),
    alpha = 0.25, size = 0.75
  ) +
  geom_sf(data = world_lcc) +
  facet_wrap(~quarter) +
  theme(legend.position = "bottom") +
  coord_sf(
    xlim = range(d$X) * 1000,
    ylim = range(d$Y) * 1000,
    crs = crs_lcc,
    expand = 0
  ) +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.direction = "horizontal",
    legend.position.inside = c(0.96, 0.07),
    legend.title = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(x = "Longitude", y = "Latitude") +
  scale_color_brewer(palette = "Dark2", direction = -1)

# Combine quarters
d <- d |>
  mutate(
    quarter_merge = ifelse(quarter %in% c(1, 2), 1.5, 3.5),
    quarter_merge = as.factor(quarter_merge)
  )

ggplot() +
  geom_point(
    data = d |> distinct(haul_id, .keep_all = TRUE),
    aes(X * 1000, Y * 1000, color = as.factor(survey)),
    alpha = 0.25, size = 0.75
  ) +
  geom_sf(data = world_lcc) +
  facet_wrap(~quarter_merge) +
  theme(legend.position = "bottom") +
  coord_sf(
    xlim = range(d$X) * 1000,
    ylim = range(d$Y) * 1000,
    crs = crs_lcc,
    expand = 0
  ) +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.direction = "horizontal",
    legend.position.inside = c(0.96, 0.07),
    legend.title = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(x = "Longitude", y = "Latitude") +
  scale_color_brewer(palette = "Dark2", direction = -1)

# Plot over time (decade)
d <- d |>
  mutate(decade = paste0(floor(year / 10) * 10, "s"))

ggplot() +
  geom_point(
    data = d |> distinct(haul_id, .keep_all = TRUE),
    aes(X * 1000, Y * 1000, color = as.factor(survey)),
    alpha = 0.25, size = 0.75
  ) +
  geom_sf(data = world_lcc) +
  facet_grid(quarter_merge ~ decade) +
  theme(legend.position = "bottom") +
  coord_sf(
    xlim = range(d$X) * 1000,
    ylim = range(d$Y) * 1000,
    crs = crs_lcc,
    expand = 0
  ) +
  theme(
    axis.text.x = element_text(angle = 0),
    legend.direction = "horizontal",
    legend.position.inside = c(0.96, 0.07),
    legend.title = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(x = "Longitude", y = "Latitude") +
  scale_color_brewer(palette = "Dark2", direction = -1)


# Loop through all species*size combos and plot presence/absence over time.
for (i in unique(d$sp_sz)) {
  dd <- filter(d, sp_sz == i) |>
    mutate(Presence = ifelse(biomass_kg == 0, "0", "1"))

  data_plot +
    geom_point(
      data = dd,
      size = 0.75,
      aes(x = X * 1000, y = Y * 1000, color = Presence)
    ) +
    scale_color_brewer(palette = "Set1") +
    geom_sf(data = world_lcc) +
    coord_sf(
      xlim = range(pred_grid$X) * 1000,
      ylim = range(pred_grid$Y) * 1000,
      crs = crs_lcc,
      expand = 0
    ) +
    ggtitle(str_replace(i, "_", " "))

  ggsave(paste0(home, "/figures/data/data_all_yrs_", dd$species[1], "_", dd$size_class[1], ".png"),
    width = 20, height = 16
  )
}

d |>
  summarise(n = n(), .by = c(quarter, survey)) |>
  ggplot(aes(quarter, n, fill = as.factor(quarter))) +
  facet_wrap(~survey, scales = "free") +
  geom_bar(stat = "identity")

d |>
  summarise(n = n(), .by = c(quarter, survey, year)) |>
  ggplot(aes(year, n, fill = as.factor(quarter))) +
  facet_wrap(~survey, scales = "free") +
  scale_fill_brewer(palette = "Dark2") +
  labs(fill = "Quarter", x = "Year") +
  coord_cartesian(expand = 0) +
  theme(legend.position = "bottom") +
  geom_bar(stat = "identity")

ggsave(paste0(home, "/figures/data/n_by_quarter.png"),
       width = 8, height = 8)
```

### Prep data for models

```{r}
#| warning: false
#| message: false

d$depth <- get.depth(depth_box, x = d$lon, y = d$lat, locator = F)$depth
d$depth <- d$depth * (-1)

# ggplot(d |>
#          distinct(haul_id, .keep_all = TRUE) |>
#          mutate(depth2 = ifelse(depth > 0, "land", "water")),
#        aes(X, Y, color = depth2)) +
#   geom_point()

d <- d |>
  tidylog::filter(depth > 1)

# Scale depth (use that for data also)
d <- d |>
  mutate(
    depth_sc = (depth - depth_mean) / depth_sd,
    depth_sc2 = depth_sc^2,
    year_f = as.factor(year),
    gear_f = as.factor(gear),
    ship_f = as.factor(ship)
  )

# Replicate grid by new quarter variable to look at seasonality
pred_grid <- pred_grid |>
  replicate_df(
    time_name = "quarter_merge",
    time_values = as.factor(c(1.5, 3.5))
  )
```

### Fit models

Fit model to all species, save predictions

```{r}
#| warning: false
#| message: false

d <- d |>
  tidylog::filter(!gear_f == "BT4P") |>
  droplevels()

tictoc::tic()

# Loop through all data and models
for (i in unique(d$sp_sz)) {
  dat <- filter(d, sp_sz == i) |>
    filter(biomass_kg < quantile(biomass_kg, probs = 0.999))

  tryCatch(
    {
      cat("Fitting:", i, "\n")

      mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 20)

      m <- sdmTMB(
        biomass_kg ~ 0 + quarter_merge + year_f + depth_sc + depth_sc2 + gear_f + (1 | ship_f),
        data = dat,
        mesh = mesh,
        # family = delta_gamma(type = "poisson-link"),
        family = tweedie(link = "log"),
        offset = log(dat$haul_dur + 5),
        spatial = "off",
        spatial_varying = ~ 0 + quarter_merge,
        spatiotemporal = "ar1",
        time = "year"
      )

      saveRDS(m, paste0(home, "/output/m_", dat$species[1], "_", dat$size_class[1], ".rds"))

      # Check residuals
      res <- simulate(m, nsim = 300, type = "mle-mvn") |>
        dharma_residuals(m, plot = FALSE)

      p_res <- res |>
        ggplot(aes(observed, expected)) +
        geom_point(color = "grey30", shape = 21, size = 0.5) +
        geom_abline(col = "tomato3", linewidth = 0.6) +
        theme(aspect.ratio = 1) +
        labs(
          x = "Observed", y = "Expected",
          title = paste("DHARMa residuals —", i)
        )

      ggsave(paste0(home, "/figures/supp/residuals_", dat$species[1], "_", dat$size_class[1], ".png"),
        width = 13, height = 14.5
      )

      # Predict
      p <- predict(
        m,
        newdata = pred_grid,
        offset = log(as.numeric(pred_grid$haul_dur) + 5),
        re_form_iid = NA
      ) |>
        mutate(
          sp_sz = i
        )

      if (m$family$family == "tweedie") {
        p <- p |> mutate(est = est)
      } else {
        p <- p |> mutate(est = est1 + est2)
      }

      saveRDS(p, paste0(home, "/output/pred_", dat$species[1], "_", dat$size_class[1], ".rds"))

      # Predict with sims to get CV
      cv_grid <- pred_grid |> filter(year_f %in% as.factor(2019:2024))

      cv_svc <- predict(
        m,
        newdata = cv_grid,
        re_form_iid = NA,
        nsim = 200, sims_var = "zeta_s"
      )

      cv_est <- predict(
        m,
        newdata = cv_grid,
        re_form_iid = NA,
        nsim = 200, sims_var = "est"
      )

      cv_grid$q1_5_zeta <- apply(cv_svc$quarter_merge1.5, 1, sd)
      cv_grid$q3_5_zeta <- apply(cv_svc$quarter_merge3.5, 1, sd)
      cv_grid$est <- apply(cv_est, 1, sd)

      cv_grid <- cv_grid |> mutate(sp_sz = i)

      saveRDS(cv_grid, paste0(home, "/output/cv_grid_", dat$species[1], "_", dat$size_class[1], ".rds"))

      # Save unique sims to illustrate stability in hotspots
      sim_grid <- pred_grid |> filter(year_f %in% as.factor(2019:2024))

      sim_grid <- sim_grid |>
        mutate(sp_sz = i) |>
        bind_cols(
          cv_est |>
            as_tibble() |>
            dplyr::select(1:30) # Just save the first 30 sims
        )

      saveRDS(sim_grid, paste0(home, "/output/sim_grid_", dat$species[1], "_", dat$size_class[1], ".rds"))

      cat("Done:", i, "\n")
    },
    error = function(e) {
      cat("ERROR:", i, "\n")
      cat(conditionMessage(e), "\n")
    }
  )
}

tictoc::tic()
```

Plotting is done in script 02

```{r}
#| warning: false
#| message: false

# Just a final check on the model convergence
# All important stuff is good. The ln_smooth_sigma is typically large,
# and might be better estimate on the full data set (all months)
list.files(
  path = paste0(home, "/output"),
  pattern = "^m_.*\\.rds$",
  full.names = TRUE
) |>
  set_names(~ tools::file_path_sans_ext(basename(.))) |>
  map(readRDS) |>
  map(sanity)
```

```{r}
knitr::knit_exit()
```

## Testing quarter specific papers

```{r}
d2 <- d |>
  tidylog::filter(!gear_f == "BT4P") |>
  filter(sp_sz == "Cod < 35 cm") |>
  droplevels()

#--- Quarter-specific models
d_1_5 <- d2 |> 
  filter(biomass_kg < quantile(biomass_kg, probs = 0.999)) |> 
  filter(quarter_merge == "1.5") |> droplevels()

d_3_5 <- d2 |> 
  filter(biomass_kg < quantile(biomass_kg, probs = 0.999)) |> 
  filter(quarter_merge == "3.5") |> droplevels()

mesh_1_5 <- make_mesh(d1_5, xy_cols = c("X", "Y"), cutoff = 20)
mesh_3_5 <- make_mesh(d3_5, xy_cols = c("X", "Y"), cutoff = 20)

m_1_5 <- sdmTMB(
  biomass_kg ~ 0 + year_f + depth_sc + depth_sc2,# + gear_f + (1 | ship_f),
  data = d_1_5,
  mesh = mesh_1_5,
  family = tweedie(link = "log"),
  offset = log(d_1_5$haul_dur + 5),
  spatial = "on",
  spatiotemporal = "iid",
  time = "year"
)

m_3_5 <- sdmTMB(
  biomass_kg ~ 0 + year_f + depth_sc + depth_sc2,# + gear_f + (1 | ship_f),
  data = d_3_5,
  mesh = mesh_3_5,
  family = tweedie(link = "log"),
  offset = log(d_3_5$haul_dur + 5),
  spatial = "on",
  spatiotemporal = "iid",
  time = "year"
)

# TODO SAVE THESE MODELS

# Predict
p_1_5 <- predict(
  m_1_5,
  newdata = pred_grid |> 
    filter(quarter_merge == "1.5") |> 
    filter(between(year, 2021, 2024)),
  #offset = log(as.numeric(pred_grid$haul_dur) + 5),
  re_form_iid = NA
)

p_3_5 <- predict(
  m_3_5,
  newdata = pred_grid |>
    filter(quarter_merge == "3.5") |> 
    filter(between(year, 2021, 2024)),
  #offset = log(as.numeric(pred_grid$haul_dur) + 5),
  re_form_iid = NA
)

# Load main model
p_combi <- readRDS(paste0(home, "/output/pred_cod_small.rds")) |> 
  filter(between(year, 2021, 2024))

#--- Plot
library(ggsidekick)

p1 <- ggplot(p_combi, aes(X, Y, fill = exp(est))) +
  geom_raster() + 
  coord_sf() +
  ggplot2::scale_fill_viridis_c(trans = "fourth_root_power") +
  facet_grid(quarter_merge~year) + 
  theme(axis.text.x = element_text(angle = 90))

p1

p_sep <- bind_rows(
  p_1_5 |>
    mutate(quarter_merge == "1.5"),
  p_3_5 |>
    mutate(quarter_merge == "3.5")
)

p2 <- ggplot(p_sep, aes(X, Y, fill = exp(est))) +
  geom_raster() + 
  coord_sf() +
  ggplot2::scale_fill_viridis_c(trans = "fourth_root_power") +
  facet_grid(quarter_merge~year) + 
  theme(axis.text.x = element_text(angle = 90))

p2

library(patchwork)
(p1 + ggtitle("Full model")) / (p2 + ggtitle("Separate model"))
```
