---
title: "Plotting overlap"
author: "Max Lindmark"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    page-layout: full
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: false
---

```{r}
#| warning: false
#| message: false

library(ggplot2)
library(sf)
library(rnaturalearth)
library(viridis)
library(tidyr)
library(dplyr)
library(purrr)
library(forcats)
library(ggsidekick)
library(raster)
library(sf)
library(patchwork)
library(mapplots)
library(scatterpie)

theme_set(theme_sleek())

home <- here::here()
crs_lcc <- 3035 # ETRS89-LAEA
# Map stuff
world <- ne_countries(scale = "medium", returnclass = "sf")
world_lcc <- world |> st_transform(crs = crs_lcc)

# Base plot

# Plot species
data_plot <- ggplot() +
  geom_sf(data = world_lcc) +
  scale_fill_viridis(
    trans = "sqrt",
    breaks = scales::pretty_breaks(n = 3)
  ) +
  guides(fill = guide_colorbar(
    title.position = "top", title.hjust = 0.5
  )) +
  theme(
    legend.position = "bottom",
    legend.key.height = unit(0.2, "cm"),
    legend.key.width = unit(0.8, "cm")
  ) +
  labs(
    x = "Longitude", y = "Latitude",
    fill = "kg per 30 min"
  )
```

### Plot predictions in space

```{r}
#| warning: false
#| message: false

# Load predictions
# see this: https://stackoverflow.com/questions/38797714/reading-multiple-rds-files
preds <- list.files(
  path = paste0(home, "/output"),
  pattern = "^pred.*\\.rds$",
  full.names = TRUE
) |>
  map(readRDS) |>
  bind_rows() |>
  mutate(
    sp_sz_code = fct_recode(
      sp_sz,
      "cod_small" = "Cod < 35 cm",
      "had_small" = "Haddock < 30 cm",
      "whg_small" = "Whiting < 27 cm",
      "cod_large" = "Cod ≥ 35 cm",
      "had_large" = "Haddock ≥ 30 cm",
      "whg_large" = "Whiting ≥ 27 cm"
    )
  ) |>
  mutate(
    est = ifelse(est > quantile(est, probs = 0.9995), quantile(est, probs = 0.9995), est),
    .by = c(sp_sz, quarter_merge)
  ) |> 
  filter(quarter_merge %in% c("1.5", "3.5"))
```

### Plot overlap (sensu NL)

```{r}
to_lcc <- function(df, lon = "lon", lat = "lat") {
  df |>
    sf::st_as_sf(
      coords = c(lon, lat),
      crs = 4326,
      remove = FALSE
    ) |>
    sf::st_transform(crs_lcc) |>
    dplyr::mutate(
      X = sf::st_coordinates(geometry)[, 1] / 1000,
      Y = sf::st_coordinates(geometry)[, 2] / 1000
    ) |>
    sf::st_drop_geometry()
}

preds_nl <- preds |> 
  mutate(ices_rect = ices.rect2(lon, lat)) |> 
  filter(between(year, 2016, 2020)) |> 
  summarise(est_natural = mean(exp(est)),
            .by = c(ices_rect, quarter_merge, sp_sz))

coords <- ices.rect(preds_nl$ices_rect)

preds_nl_wide <- preds_nl |> 
  mutate(lon = coords$lon,
         lat = coords$lat) |> 
  pivot_wider(names_from = sp_sz, values_from = est_natural)

preds_nl_wide <- preds_nl_wide |> 
  mutate(sum_juv = `Cod < 35 cm` + `Haddock < 30 cm` + `Whiting < 27 cm`,
         juv_cod_prop = `Cod < 35 cm` / sum_juv,
         juv_had_prop = `Haddock < 30 cm` / sum_juv,
         juv_whi_prop = `Whiting < 27 cm` / sum_juv) |> 
  to_lcc(lon = "lon", lat = "lat")

spp <- c("Cod < 35 cm", "Haddock < 30 cm", "Whiting < 27 cm")
r_max <- 25  # km

juv_cols <- c(
  "Cod < 35 cm"     = "#F6EBCA",
  "Haddock < 30 cm" = "#98652B",
  "Whiting < 27 cm" = "#1D746E"
)

pie_dat <- preds_nl_wide |>
  filter(sum_juv > 0) |>
  mutate(x_m = X * 1000,
         y_m = Y * 1000,
         r = 35000 * (sum_juv / max(sum_juv))^0.2)

pie_dat <- pie_dat |>
  mutate(all_5 = juv_cod_prop > 0.05 & juv_had_prop > 0.05 & juv_whi_prop > 0.05)

ggplot() +
  geom_sf(data = world_lcc) +
  geom_scatterpie(
    data = pie_dat,
    alpha = 0.8,
    cols = spp,
    color = NA,
    aes(x = x_m, y = y_m, r = r)
  ) +
  ggforce::geom_circle(
    data = filter(pie_dat, all_5),
    aes(x0 = x_m, y0 = y_m, r = r),
    color = "tomato",
    linewidth = 0.4,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = juv_cols) +
  facet_wrap(~quarter_merge) +
  coord_sf(
    xlim = range(pie_dat$x_m) + c(-1, 1) * 50000,
    ylim = range(pie_dat$y_m) + c(-1, 1) * 50000,
    crs = crs_lcc,
    expand = 0
  ) +
  labs(x = "Longitude", y = "Latitude", fill = NULL) +
  theme(
    legend.position = "bottom",
    legend.key.height = unit(0.4, "cm"),
    legend.key.width = unit(0.4, "cm")
  )
#ggsave(paste0(home, "/figures/test_pie.png"), width = 19, height = 10, units = "cm")
```

### Pairwise spatial overlap (Bhattacharyya's coefficient)

Symmetric proportional co-occurrence between each pair of species: each
species' density surface is normalised to proportions (sums to 1 across the
grid), and BC = sum_i sqrt(p1i * p2i). Identical distributions give BC = 1,
completely disjoint distributions give BC = 0. The per-cell term
sqrt(p1i * p2i) is the map below; it sums to the reported BC, so the map and
the number are the same quantity -- cells are bright only where *both*
species hold a meaningful share of their own population, regardless of which
is more abundant overall (Carroll et al. 2019).

Same setup as `02-plot-ternary-tricolore.qmd`: a pooled map for the last 5
years, and a year-faceted map for the last 20 years, saved to
`figures/bc/`.

```{r}
#| warning: false
#| message: false

dir.create(paste0(home, "/figures/bc"), showWarnings = FALSE, recursive = TRUE)

sp_labels <- c(cod = "Cod", had = "Haddock", whg = "Whiting")
pairs <- list(c("cod", "had"), c("cod", "whg"), c("had", "whg"))

# by_year = TRUE keeps year as a grouping variable (proportions normalised
# WITHIN each year) instead of averaging density across years into one map
# first.
build_bc <- function(preds, years, by_year = FALSE) {
  group_vars <- c(if (by_year) "year", "X", "Y", "species", "size_class")
  d <- preds |>
    filter(year %in% years) |>
    mutate(
      species = sub("_.*", "", sp_sz_code),
      size_class = ifelse(grepl("large$", sp_sz_code), "Adult", "Juvenile"),
      dens = exp(est)
    ) |>
    summarise(dens = mean(dens), .by = all_of(group_vars))

  norm_vars <- c(if (by_year) "year", "size_class", "species")
  d <- d |> mutate(p = dens / sum(dens), .by = all_of(norm_vars))

  wide_id <- c(if (by_year) "year", "X", "Y", "size_class")
  d_wide <- d |>
    dplyr::select(all_of(c(wide_id, "species", "p"))) |>
    pivot_wider(names_from = species, values_from = p, names_prefix = "p_", values_fill = 0)

  bc_long <- map_dfr(pairs, function(sp) {
    a <- paste0("p_", sp[1]); b <- paste0("p_", sp[2])
    d_wide |>
      mutate(
        pair = paste(sp_labels[sp[1]], "–", sp_labels[sp[2]]),
        term = sqrt(.data[[a]] * .data[[b]])
      ) |>
      dplyr::select(all_of(c(wide_id, "pair", "term")))
  })

  sum_vars <- c(if (by_year) "year", "size_class", "pair")
  bc_summary <- bc_long |> summarise(BC = sum(term), .by = all_of(sum_vars))

  list(long = bc_long, summary = bc_summary)
}
```

```{r}
#| warning: false
#| message: false

plot_bc_single <- function(bc, sc, title, out_file) {
  d_summary <- bc$summary |> filter(size_class == sc)

  # BC value folded straight into the facet strip text (e.g. "Cod -
  # Haddock (BC = 0.74)") instead of a floating on-map label -- the strip
  # always has a plain, consistent background, so there's no risk of it
  # landing on the wrong part of the raster and becoming unreadable.
  d <- bc$long |>
    filter(size_class == sc) |>
    left_join(d_summary, by = c("size_class", "pair")) |>
    mutate(pair_lab = paste0(pair, " (BC = ", sprintf("%.2f", BC), ")"))

  xlim <- range(d$X) * 1000 + c(-1, 1) * 50000
  ylim <- range(d$Y) * 1000 + c(-1, 1) * 50000

  p <- ggplot() +
    geom_raster(data = d, aes(X * 1000, Y * 1000, fill = term)) +
    geom_sf(data = world_lcc, fill = "grey90", colour = "grey60", linewidth = 0.2) +
    scale_fill_viridis(
      option = "magma", trans = "sqrt",
      breaks = scales::pretty_breaks(n = 3)
    ) +
    facet_wrap(~pair_lab, nrow = 1) +
    coord_sf(xlim = xlim, ylim = ylim, crs = crs_lcc, expand = 0) +
    labs(x = "Longitude", y = "Latitude", fill = "sqrt(p1 x p2)", title = title) +
    theme_sleek() +
    theme(
      legend.position = "bottom",
      legend.key.height = unit(0.2, "cm"),
      legend.key.width = unit(0.8, "cm"),
      strip.text = element_text(size = 8)
    )

  ggsave(out_file, p, width = 21, height = 8, units = "cm", dpi = 300)
  p
}

plot_bc_facets <- function(bc, sc, title, out_file) {
  d <- bc$long |> filter(size_class == sc)

  xlim <- range(d$X) * 1000 + c(-1, 1) * 50000
  ylim <- range(d$Y) * 1000 + c(-1, 1) * 50000

  p <- ggplot() +
    geom_raster(data = d, aes(X * 1000, Y * 1000, fill = term)) +
    geom_sf(data = world_lcc, fill = "grey90", colour = "grey60", linewidth = 0.2) +
    scale_fill_viridis(
      option = "magma", trans = "sqrt",
      breaks = scales::pretty_breaks(n = 3)
    ) +
    facet_grid(pair ~ year) +
    coord_sf(xlim = xlim, ylim = ylim, crs = crs_lcc, expand = 0) +
    labs(x = "Longitude", y = "Latitude", fill = "sqrt(p1 x p2)", title = title) +
    theme_sleek() +
    theme(
      legend.position = "bottom",
      legend.key.height = unit(0.2, "cm"),
      legend.key.width = unit(0.8, "cm"),
      axis.text = element_text(size = 5),
      strip.text = element_text(size = 7)
    )

  ggsave(out_file, p, width = 40, height = 14, units = "cm", dpi = 180)
  p
}
```

```{r}
#| warning: false
#| message: false

ref_year <- max(preds$year)
years_5 <- (ref_year - 4):ref_year
years_20 <- (ref_year - 19):ref_year
yr5_lab <- paste0(min(years_5), "-", max(years_5))
yr20_lab <- paste0(min(years_20), "-", max(years_20))

bc_pooled <- build_bc(preds, years_5)
bc_facets <- build_bc(preds, years_20, by_year = TRUE)

plot_bc_single(
  bc_pooled, "Juvenile",
  paste("Juvenile spatial overlap (Bhattacharyya's coefficient),", yr5_lab),
  paste0(home, "/figures/bc/juvenile_", yr5_lab, ".png")
)
plot_bc_single(
  bc_pooled, "Adult",
  paste("Adult spatial overlap (Bhattacharyya's coefficient),", yr5_lab),
  paste0(home, "/figures/bc/adult_", yr5_lab, ".png")
)

plot_bc_facets(
  bc_facets, "Juvenile",
  "Juvenile spatial overlap (Bhattacharyya's coefficient)",
  paste0(home, "/figures/bc/juvenile_", yr20_lab, ".png")
)
plot_bc_facets(
  bc_facets, "Adult",
  "Adult spatial overlap (Bhattacharyya's coefficient)",
  paste0(home, "/figures/bc/adult_", yr20_lab, ".png")
)
```
